Practical Parametricity for GADTs

### 2020-10-01 20-gadts/inf-rules.tex
added Lan-intro-interp to inf-rules.tex

\[\begin{array}{c}
\AXC{$\Gamma; \Phi \vdash F$}
\AXC{$\Gamma; \Phi \vdash G$}
\AXC{$\Gamma; \Phi \,|\, \Delta, x : F \vdash t: G$} 
\TIC{$\Gamma; \emptyset
  \,|\, \Delta \vdash L_\Phi x.t : \Nat^\Phi \,F \,G$}
\DisplayProof
\\\\
\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
\AXC{$\Gamma; \emptyset
  \,|\, \Delta \vdash t : \Nat^{\ol\psi} \,F \,G$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\psi :=_{\ol\beta} K}]$}
\TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
  G[\overline{\psi :=_{\ol\beta} K}]$}
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi, \ol{\phi} \vdash H$}
\AXC{$\ol{\Gamma; \Phi,\ol{\beta} \vdash F}$}
\AXC{$\ol{\Gamma; \Phi,\ol{\beta} \vdash
    G}$}
\TIC{$\Gamma; \emptyset
  ~|~\emptyset
  \vdash \map^{\ol{F},\ol{G}}_H :
  \Nat^\emptyset\;(\ol{\Nat^{\Phi,\ol{\beta}}\,F\,G})\;
  (\Nat^\Phi\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
      :=_{\ol{\beta}} G}])$} 
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi, \phi, \ol{\alpha} \vdash H$}
\UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
  \Nat^{\Phi,\ol{\beta}} H[\phi :=_{\ol{\beta}} (\mu
    \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
  \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
\DisplayProof
\\\\
\AXC{$\Gamma; \phi,\Phi,\ol{\alpha} \vdash H$}
\AXC{$\Gamma; \Phi, \ol{\beta} \vdash F$}
\BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
  \Nat^\emptyset\; (\Nat^{\Phi,\ol{\beta}}\,H[\phi
    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\Phi,\ol{\beta}}
    \,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
%\AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
\BIC{$\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F} :
\Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\,  F)\ol{K}$}  
\DisplayProof
\\\\
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
  F\;G[\ol{\beta := K}]$}
%\AXC{$\ol{\Gamma;\Phi \vdash B}$}
%\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\ol K} F)\,\ol B $}
\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta\; G$}
\DisplayProof
\end{array}\]
\end{dfn}

@@

@@

\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
: \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\, F)\ol{K}}\rho
& = & \vspace*{0.05in}\\
\relsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F} :
  \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\,
  F)\ol{K}}\rho\,d & = & \vspace*{0.05in} \\

@@

\new{In the fourth-to-last clause of Definition~\ref{def:set-interp},...}

@@

\new{In the final last clause of Definition~\ref{def:set-interp},...}

@@

\subsection{$\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
: \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\, F)\ol{K}}$ is
  well-defined}

\subsection{$\relsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
: \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\, F)\ol{K}}$ is
  well-defined}

\subsection{$\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol
      K}_F \eta : Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
    F)\ol \beta\; G}$ is well-defined} We must show that, for every
$\rho : \setenv$ and $d : \setsem{\Gamma; \emptyset \vdash \Delta}
\rho$, \[\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \partial^{G,
    \ol K}_F \eta : \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
  F)\ol \beta\; G} \rho \, d\] has the correct type and satisfies the
condition required for it to be in $\setsem{\Gamma; \emptyset \vdash
  \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta\; G}
\rho$. For the former we observe that each component
$(\mu_{\ol{N}})_{\ol{B}}$ of $\mu_{\ol{N}}$ has type
  \[\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho
  {[\ol{\alpha := B} ]}}} \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho
        {[\Phi, \ol \alpha := \ol N, \ol B]} \to \setsem{\Gamma; \Phi,
          \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \ol B]\]
        Each component of $(\mu_{\ol{N}})_{\ol{B}} |_{\setsem{\Gamma;
            \Phi, \ol\beta \vdash (\Lan^{\ol\alpha}_{\ol K} F)\ol
            \beta } \rho[\Phi, \ol \beta := \ol N, \ol B]}$ therefore
        has the correct type to be the component at $\ol N$ and $\ol
        B$ of a natural transformation in $\setsem{\Gamma; \emptyset
          \vdash \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
          F)\ol \beta\; G} \rho$.

@@

\subsection{$\relsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol
      K}_F \eta : Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
    F)\ol \beta\; G}$ is well-defined}

>>>>

\[\begin{array}{c}
\AXC{$\Gamma; \Phi \vdash F$}
\AXC{$\Gamma; \Phi \vdash G$}
\AXC{$\Gamma; \Phi \,|\, \Delta, x : F \vdash t: G$} 
\TIC{$\Gamma; \emptyset
  \,|\, \Delta \vdash L_\Phi x.t : \Nat^\Phi \,F \,G$}
\DisplayProof
\\\\
\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
\AXC{$\Gamma; \emptyset
  \,|\, \Delta \vdash t : \Nat^{\ol\psi} \,F \,G$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\psi :=_{\ol\beta} K}]$}
\TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
  G[\overline{\psi :=_{\ol\beta} K}]$}
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi, \ol{\phi} \vdash H$}
\AXC{$\ol{\Gamma; \Phi,\ol{\beta} \vdash F}$}
\AXC{$\ol{\Gamma; \Phi,\ol{\beta} \vdash
    G}$}
\TIC{$\Gamma; \emptyset
  ~|~\emptyset
  \vdash \map^{\ol{F},\ol{G}}_H :
  \Nat^\emptyset\;(\ol{\Nat^{\Phi,\ol{\beta}}\,F\,G})\;
  (\Nat^\Phi\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
      :=_{\ol{\beta}} G}])$} 
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi, \phi, \ol{\alpha} \vdash H$}
\UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
  \Nat^{\Phi,\ol{\beta}} H[\phi :=_{\ol{\beta}} (\mu
    \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
  \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
\DisplayProof
\\\\
\AXC{$\Gamma; \phi,\Phi,\ol{\alpha} \vdash H$}
\AXC{$\Gamma; \Phi, \ol{\beta} \vdash F$}
\BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
  \Nat^\emptyset\; (\Nat^{\Phi,\ol{\beta}}\,H[\phi
    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\Phi,\ol{\beta}}
    \,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
\AXC{$\ol{\Gamma; \Phi \vdash A}$}
\AXC{$\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]$}
\QIC{$\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t : (\Lan^{\ol\alpha}_{\ol{K}}\,  F)\, \ol{K[\ol{\alpha := A}]}$}
\DisplayProof
\\\\
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
  F\;G[\ol{\beta := K}]$}
%\AXC{$\ol{\Gamma;\Phi \vdash B}$}
%\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\ol K} F)\,\ol B $}
\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta\; G$}
\DisplayProof
\end{array}\]
\end{dfn}

@@

\new{Note that we cannot have
\[
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$} \AXC{$\ol{\Gamma; \ol{\alpha}
    \vdash K}$} \BIC{$\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol
    K,F} : \Nat^{\Phi,\ol\alpha}\,F\, (\Lan^{\ol\alpha}_{\ol K}\,
  F)\ol{K}$} \DisplayProof\] We cannot prove the additional property
needed to define an object semantics for this rule because we can only
introduce components of the natural transformation at representable
types, as in the rule above. We cannot produce the entire natural
transformation required in the conclusion of the alternative rule
here. This makes sense because not all parts of the functors
underlying GADTs are representable in syntax, which in turn means that
not all components of the associated natural transformations are
representable in syntax.}

@@

\setsem{\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
  (\Lan^{\ol\alpha}_{\ol{K}}\,  F)\, \ol{K[\ol{\alpha := A}]}}\rho\;d
& = & = \iota_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho},
  \ol{\id_{\oCPOsem{\Gamma; \Phi \vdash K[\ol{\alpha := A}]}\rho}}}
\oCPOsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\, 
d \vspace*{0.05in}\\
\relsem{\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
  (\Lan^{\ol\alpha}_{\ol{K}}\,  F)\, \ol{K[\ol{\alpha := A}]}}\rho\;d
& = & \vspace*{0.05in} \\

@@

\new{In the fourth-to-last clause of Definition~\ref{def:set-interp},
  $\iota_{Z, f}$ is the injection into
\[\colim{\ol{Z : \Set_0},\, \ol{f : \oCPOsem{\Gamma; \ol\alpha \vdash K}
    \rho[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
      K[\ol{\alpha := A}]} \rho}}{\,\oCPOsem{\Gamma; \Phi, \ol\alpha
    \vdash F} \rho[\ol{\alpha := Z}]}\]

}

@@

\new{In the final clause of Definition~\ref{def:set-interp},...}

@@

\subsection{Object interpretations of $\Lan$-introduction
  are well-defined}

We must show that, for every $\rho : \setenv$ and $d :
\oCPOsem{\Gamma;\Phi \vdash \Delta}\rho$,
\[\oCPOsem{\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
  (\Lan^{\ol\alpha}_{\ol{K}}\, F)\, \ol{K[\ol{\alpha := A}]}}\rho\, d
\in \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol\alpha}_{\ol{K}}\, F)\,
  \ol{K[\ol{\alpha := A}]}}\rho\] i.e., that there exist $\ol{Z :
  \oCPO_0}$, $t_1', t_2' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash
  F}\rho[\ol{\alpha := Z}]$, and
\[\ol{f : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\Eq_{\rho}[\ol{\alpha
      := \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash K[\ol{\alpha :=
        A}]}\Eq_{\rho}} \]
such that
\[\begin{array}{rcl}
\iota_{\ol{Z},\ol{\pi_1 f}}t_1' &=& \iota_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho}, \ol{\id_{\oCPOsem{\Gamma; \Phi \vdash K[\ol{\alpha := A}]}\rho}}}\,
\oCPOsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\, d \\[2ex]
\iota_{\ol{Z}, \ol{\pi_2 f}}t_2' &=& \iota_{\ol{\oCPOsem{\Gamma;\Phi \vdash A}\rho}, \ol{\id_{\oCPOsem{\Gamma; \Phi \vdash K[\ol{\alpha := A}]}\rho}}}\,
\oCPOsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\,
d\\[2ex]
(t_1',t_2') &\in& \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash
  F}\Eq_{\rho}[\ol{\alpha := \Eq_Z}] \;=\; \oCPORsem{\Gamma;
  \Phi, \ol\alpha \vdash F}\Eq_{\rho[\ol{\alpha := Z}]}
\end{array}\]
Let $\ol{Z = \oCPOsem{\Gamma; \Phi \vdash A} \rho}$, $t_1' = t_2' =
\oCPOsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\,
d$, and $\ol{f = \id_{\oCPORsem{\Gamma; \ol\alpha \vdash K[\ol{\alpha
          := A}]}\Eq_{\rho}}}$.  Then~\eqref{eq:subs-var} gives that
$t_1', t_2' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash
  F}\rho[\ol{\alpha := Z}]$ as required, as well as that
$\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\Eq_{\rho}[\ol{\alpha :=
    \Eq_{\oCPOsem{\Gamma; \Phi \vdash A} \rho}}] = \oCPORsem{\Gamma;
  \Phi \vdash K[\ol{\alpha := A}]}\Eq_{\rho}$. In light of the latter
we take $f$ to be
\[\ol{ \id_{\oCPORsem{\Gamma; \ol\alpha \vdash K[\ol{\alpha :=
          A}]}\Eq_{\rho}} : \oCPORsem{\Gamma; \ol{\alpha} \vdash
    K}\Eq_{\rho}[\ol{\alpha := \Eq_{\oCPOsem{\Gamma; \Phi \vdash A}
        \rho}}] \to \oCPORsem{\Gamma; \Phi \vdash K[\ol{\alpha :=
        A}]}\Eq_{\rho} }\] Then the first two properties above hold by
construction, and the third holds because $t_1' = t_2'$ and because
$\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho[\ol{\alpha :=
      Z}]} = \Eq_{\setsem{\Gamma; \Phi, \ol\alpha \vdash
    F}\rho[\ol{\alpha := Z}]}$ by the IEL.

\subsection{Relational interpretations of $\Lan$-introduction are
  well-defined}

\subsection{Object interpretations of $\Lan$-elimination are
  well-defined} We must show that, for every 
$\rho : \setenv$ and $d : \setsem{\Gamma; \emptyset \vdash \Delta}
\rho$, \[\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \partial^{G,
    \ol K}_F \eta : \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
  F)\ol \beta\; G} \rho \, d\] has the correct type and satisfies the
condition required for it to be in $\setsem{\Gamma; \emptyset \vdash
  \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K} F)\ol \beta\; G}
\rho$. For the former we observe that each component
$(\mu_{\ol{N}})_{\ol{B}}$ of $\mu_{\ol{N}}$ has type
  \[\Lan_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho
  {[\ol{\alpha := B} ]}}} \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho
        {[\Phi, \ol \alpha := \ol N, \ol B]} \to \setsem{\Gamma; \Phi,
          \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \ol B]\]
        Each component of $(\mu_{\ol{N}})_{\ol{B}} |_{\setsem{\Gamma;
            \Phi, \ol\beta \vdash (\Lan^{\ol\alpha}_{\ol K} F)\ol
            \beta } \rho[\Phi, \ol \beta := \ol N, \ol B]}$ therefore
        has the correct type to be the component at $\ol N$ and $\ol
        B$ of a natural transformation in $\setsem{\Gamma; \emptyset
          \vdash \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}
          F)\ol \beta\; G} \rho$.

@@

\subsection{Relational interpretations of $\Lan$-elimination are
  well-defined}
